---
title: "lab19"
author: "Nicholas Pacia"
format: pdf
---

# Is Pertussis Increasing

CDC tracks reported Pertussis cases.

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
#install.packages("datapasta")
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
head(cdc)
```
```{r}
library(ggplot2)
pertussis_plot <- ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases by year (1922-2019)", x = "Year", y = "Number of cases")
pertussis_plot
```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
pertussis_plot <- pertussis_plot + geom_vline(xintercept = 1946, color = "blue",
                            linetype = "dashed" ) +
  geom_vline(xintercept = 1996, color = "red", linetype = "dashed")+
  annotate("text", x=1948, y = 250000, label = "wP", color = "blue")+
  annotate("text", 1998, 250000, label = "aP", color = "red")
pertussis_plot
```


>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the aP vaccine, there was an increase in pertussis cases. This can be from an increase in people not taking vaccinations, evolution and adaptation from the bacteria, and the loss of immunity from the earlier generations.

# Getting Data from CMI-PB

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject)
```

>Q4. How may aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject["infancy_vac"])
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

# Working with Dates

```{r}
library(lubridate)
today()
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
library(dplyr)

print("wp")
wp <- subject %>% filter(infancy_vac == "wP")
wp_age <- round(summary(time_length(today() - ymd(wp$year_of_birth), "years")))
wp_age

print("ap")
ap <- subject %>% filter(infancy_vac == "aP")
ap_age <- round(summary(time_length(today() - ymd(ap$year_of_birth), "years")))
ap_age

t.test(wp$age, ap$age)
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
boost <- (time_length(ymd(subject$date_of_boost) - ymd(subject$year_of_birth), "years"))
head(boost)
```

>Q9. With the help of a faceted boxplot (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2)
```
Yes these groups are significantly different. aP vaccines were all given most recently (younger than the age of 20's), while wP vaccines were given to very person above this age.

# Conjoining tables

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
```


>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```
Visit 8 has significantly fewer entries than the other visits. 

# Examing IgG1 Ab titer levels

```{r}
#remove visit 8 with too few entries
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2) + 
  labs(x="ab_titer")
```

>Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

FIM2/3, PRN, FHA, and DT seem to show differences in levels over time, because they resond to vaccine more than the other antigens.



```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).

```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "Measles antigen levels per visit")
```
```{r}
filter(ig1, antigen== "FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "FIM2/3 antigen levels per visit")
```

>Q16. What do you notice about these two antigens time course and the FIM2/3 data in particular?

FIM2/3 levels increase over time and exceed the levels of Measles. FIM2/3 seems to peak around visit 5 where it begins to decline. The aP vaccine maintains higher levels of FIM2/3 after visit 4, which is opposite for Measles, but the overall trend is similar between aP and wP vaccines.

>Q17. Do you see any clear difference in aP vs. wP responses?

For measles, wP has a slightly higher antigen count than the aP vaccine. For FIM2/3, aP had lower antigen count early on but had a larger count for the later visits, so aP vaccine had less long term efficacy.







