---
title: "Lab 11"
author: "Nicholas Pacia"
format: pdf
---

# Section 4 Population Scale Analysis

First download the file fromm https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt into the project directory. Then read the file.
```{r}
url <- "rs8067378_ENSG00000172057.6.txt"
data <- read.table(url)
head(data)
```

>Q13 What is the sample size for each genotype and their median expression levels?

```{r}
nrow(data) #total size of sample
table(data$geno) #sample size of each genotype
```
```{r}
#separating the genotypes into individual vectors to find median expression
aa <- data[data$geno == "A/A",]
ag <- data[data$geno == "A/G",]
gg <- data[data$geno == "G/G",]
```
```{r}
#two different ways of getting median expression if already separated
median(aa$exp)
median(data[data$geno == "A/G", "exp"])
median(gg$exp)
```


```{r}
library(ggplot2)
```
```{r}
ggplot(data, aes(geno,exp, fill = geno)) +
  geom_boxplot(notch = T) 
```

>Q14 What could you infer from the realtive expression value between A/A and G/G displayed in this plot? does the SNP effect expression of ORMDL3

A/A has a higher expression than G/G so the G/G genotype has reduced expression of this gene. Depending on the SNP, the expression of ORMDL3 changes; it is higher for A's and lower for G's.

